---
title: "The Route 508 Streetcar and Congestion on Toronto's Shoreline, 2011 - 2020"
subtitle: "Examining Traffic Patterns Before and After Public Transit Improvements"
author: 
  - Maria Mangru
thanks: "Code and data are available at: https://github.com/MariaMangru/Traffic-Congestion-on-Toronto-s-Shoreline."
date: today
date-format: long
abstract: "This study investigates the impact of enhanced public transit infrastructure on car congestion. It specifically focuses on the reintroduction of the Route 508 Lake Shore streetcar between 2011 and 2020. The data, sourced from the City of Toronto's Transportation Services Division, includes detailed traffic volumes for various modes at intersections across the city. The method of analysis includes comparing the negative binomial regression models of daily car traffic before and after transit upgrades. The results highlighted a significant reduction in the relationship between bus traffic and car congestion. Additionally, it was observed that car traffic decreased over time. These findings highlight the usefulness of public transit improvements in reducing urban congestion."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r, echo=FALSE, message=FALSE}
#| include: false
#| warning: false
#| message: false

library(dplyr)
library(ggplot2)
library(tidyr)
library(readr)
library(here)
library(knitr)
library(kableExtra)
library(stringr)
library(broom)
library(MASS) 
library(car)
```

# Introduction

Traffic congestion represents a significant challenge for urban cities. It is a non-productive activity which negatively affects economic efficiency, environmental sustainability and overall quality of life. To address this challenge, many cities like Toronto have undergone improvements to public transit infrastructure with the aim of reducing car dependency and mitigate traffic congestion. However, it is still highly debatable whether or not such improvements lead to a decrease in car congestion.

This study specifically investigates the 508 Lake Shore Streetcar route in Toronto. It utilizes two datasets provided by the City of Toronto's Transportation Services Division. These datasets capture traffic volumes for various modes at city intersections. To analyze this data, @citeR was used along with several R packages. Through the use of a negative binomial regression, we can assess daily car traffic relative to bus traffic, pedestrian, bicycle flows and a time trend. Models were created to investigate traffic patterns before and after public transit improvements.

The estimand of this study is quantitatively evaluated through the change in the relationship between daily bus traffic and daily car traffic, as well as the evolution of car traffic trends over time. The findings from this study indicates a significant reduction in the relationship between bus traffic and car congestion, following the transit improvements. Additionally, a reversal from an increasing to a decreasing trend in car traffic over time was noted.

This research begins with an in-depth exploration of the dataset and the analytical methods employed, highlighting the use of @citeR for data analysis. It then goes on to explain the results of the regression model, comparing and contrasting traffic patterns prior to and following the installation of transportation upgrades. The discussion interprets these findings within the broader framework of managing urban congestion and the importance of public transit. Through this approach, I hope to illustrate the important role of public transit improvements in combating urban congestion, as well as its significance for promoting sustainable urban mobility and improving the standard of living in urban areas.

# Data {#sec-data}

## Data Source

This research relies on traffic volume data sets from the City of Toronto's Transportation Services Division. It accessible through the Toronto OpenData portal and is titled "Traffic Volumes at Intersections for All Modes" which is free and accessible for public use. The information collected in the dataset are of two main types:

1.  Automatic Traffic Recorder Counts (ATR): These are segment-level volumes which capture the total number of vehicles, cyclists, or pedestrians moving in a specific direction on a street.

2.  Turning Movement Counts (TMCs): These detail the volume observed at each leg of an intersection, including the turning movement by mode (car, truck, bus, pedestrian, cyclist, other).

The data set covers various intersections across Toronto, providing a detailed view of the city's traffic dynamics. It includes data spanning 2010-2019 and 2020-2024 which was combined into one data set. The data set is segmented by direction of approach, turning movement, and mode, in 15-minute intervals. This level of detail allows for a thorough investigation of traffic patterns, which is especially important for analyzing how changes to public transport affect traffic congestion on the 508 Lake Shore route.

The data set was cleaned and prepared, with yearly data files combined, relevant date periods filtered, and traffic volumes aggregated to get daily totals for every mode of transportation. This pre-processing step made sure the data was suitable for analysis, with a particular emphasis on the time frame prior to and following the 508 Lake Shore route's streetcar upgrades.

## Variables of Interest

In addition to a temporal trend variables, specific variables were chosen for this analysis, with an emphasis on the dynamics of car traffic in relation to bus, pedestrian, and bicycle traffic volumes:

-   Daily Car Traffic (`daily_cars`): Total volume of car traffic recorded at selected intersections along the 508 Route.
-   Daily Bus Traffic (`daily_bus`), Pedestrian Traffic (`daily_peds`) and Bicycle Traffic (`daily_bike`): Bus, pedestrian and bike traffic volumes used to indicate the presence and frequency of non personal car usage.
-   Time Trend (time_trend): A constructed variable to analyze traffic pattern changes over time.

Using @citeR, the dataset was processed by using tools like `ggplot2` for visualization, `dplyr` for data manipulation, and `lubridate` for date management. These packages made it easier to clean, aggregate, and analyze traffic volumes, which helped to provide a more complex picture of traffic patterns and flow.

```{r summary-stats-table, echo=FALSE, message=FALSE, warning=FALSE}
# Load the cleaned data
cleaned_data <- read_csv(here("data", "analysis_data", "cleaned-data.csv"))

# Generate summary statistics for traffic data
summary_stats <- summarise(cleaned_data,
  Mean_Cars = mean(daily_cars, na.rm = TRUE),
  Median_Cars = median(daily_cars, na.rm = TRUE),
  Mean_Buses = mean(daily_bus, na.rm = TRUE),
  Median_Buses = median(daily_bus, na.rm = TRUE),
  Mean_Pedestrians = mean(daily_peds, na.rm = TRUE),
  Median_Pedestrians = median(daily_peds, na.rm = TRUE),
  Mean_Bicycles = mean(daily_bike, na.rm = TRUE),
  Median_Bicycles = median(daily_bike, na.rm = TRUE)
)

# Create a data frame for the kable function
summary_stats_df <- data.frame(
  Transport_Mode = c("Daily Car Traffic", "Daily Bus Traffic", "Daily Pedestrian Traffic", "Daily Bicycle Traffic"),
  Mean = c(summary_stats$Mean_Cars, summary_stats$Mean_Buses, summary_stats$Mean_Pedestrians, summary_stats$Mean_Bicycles),
  Median = c(summary_stats$Median_Cars, summary_stats$Median_Buses, summary_stats$Median_Pedestrians, summary_stats$Median_Bicycles)
)

# Display summary statistics
kable(summary_stats_df, format = "html", 
      col.names = c("Transport Mode", "Mean", "Median"),
      caption = "Summary statistics table for daily traffic volume",
      table.attr = 'style="width:100%;"') 
```

The summary table above provides a quantitative comparison across transportation modes by condensing the traffic data for each mode into average and median statistics. The data reveals that car traffic surpasses other modes in volume, with the mean and median values for daily car traffic being the highest.

```{r line-graph-cars, fig.cap="Daily Car Traffic Over Time. This graph displays the trend in number of cars within the area of interest between December 2011 to December 2014 and November 2018 to February 2020.",fig.width=8, fig.height=4, echo=FALSE, message=FALSE, warning=FALSE}

# Visualizing daily car traffic over time
line_graph_car <- cleaned_data %>%
  ggplot(aes(x = count_date, y = daily_cars)) +
  geom_smooth(se = FALSE) +  
  labs(title = "Daily Cars Traffic Over Time",
       x = "Date",
       y = "Total Cars Traffic") +
  theme_minimal()

# Display car traffic plot
line_graph_car

```

The daily variation in car traffic from 2011 to 2020 is shown in Figure 1. The traffic volume is depicted on a smooth line graph that first increases and peaks about 2016, after which it sharply declines. The fall that follows may point to the benefits of the Route 508 Lake Shore streetcar improvements on easing traffic congestion. The peak denotes a period of increased traffic.

```{r line-graph-bus, fig.cap="Daily Bus Traffic Over Time. This graph displays the trend in number of buses within the area of interest between December 2011 to December 2014 and November 2018 to February 2020.", fig.width=8, fig.height=4, echo=FALSE, message=FALSE, warning=FALSE}

# Visualizing daily bus traffic over time
line_graph_buses <- cleaned_data %>%
  ggplot(aes(x = count_date, y = daily_bus)) +
  geom_smooth(se = FALSE) +
  labs(title = "Daily Bus Traffic Over Time",
       x = "Date",
       y = "Total Bus Traffic") +
  theme_minimal()

# Display bus traffic plot
line_graph_buses

```

Figure 2 depicts the daily patterns in bus traffic, which follow a similar pattern to that of car traffic. It is a smooth line graph that rises to a high around in 2016, then declines. This pattern might be an indication of how bus utilization has changed, which could be related to the streetcar route's service modifications and other improvements to public transportation.

```{r line-graph-pedestrian, fig.cap="Daily Pedestrian Traffic Over Time. This graph displays the trend in number of pedestrian within the area of interest between December 2011 to December 2014 and November 2018 to February 2020.", fig.width=8, fig.height=4, echo=FALSE, message=FALSE, warning=FALSE}

# Visualizing daily pedestrian traffic over time
line_graph_pedestrians <- cleaned_data %>%
  ggplot(aes(x = count_date, y = daily_peds)) +
  geom_smooth(se = FALSE) +
  labs(title = "Daily Pedestrian Traffic Over Time",
       x = "Date",
       y = "Total Pedestrian Traffic") +
  theme_minimal()

# Display pedestrian traffic plot
line_graph_pedestrians

```

Variations in pedestrian traffic over is shown in Figure 3's line graph. The middle of the time period has a noteworthy decrease, which is followed by a rise and another decline. This could be attributed to a number of things, such as changes in walkability, urban development, and perhaps the availability of better public transportation alternatives that encourage walking as a mode of transportation.

```{r line-graph-bike, fig.cap="Daily Bike Traffic Over Time. This graph displays the trend in number of bikes within the area of interest between December 2011 to December 2014 and November 2018 to February 2020.", fig.width=8, fig.height=4, echo=FALSE, message=FALSE, warning=FALSE}

# Visualizing daily bike traffic over time
line_graph_cyclists <- cleaned_data %>%
  ggplot(aes(x = count_date, y = daily_bike)) +
  geom_smooth(se = FALSE) +
  labs(title = "Daily Cyclist Traffic Over Time",
       x = "Date",
       y = "Total Cyclist Traffic") +
  theme_minimal()

# Display cyclist traffic plot
line_graph_cyclists

```

The pattern of bicycle traffic over time is depicted in Figure 4, where a notable rise and fall are shown on the line graph, indicating huge variations. The variations may be attributed to changing public opinions on riding, the implementation of bike infrastructure, or seasonal variations in weather patterns, which may be exacerbated by changes made to the transit system.

```{r line-graph-time-trend, fig.cap="Daily Car Traffic Over Time. This graph displays the general trend in number of cars from the beginning of data collection period.", fig.width=8, fig.height=4, echo=FALSE, message=FALSE, warning=FALSE}

# Visualizing car traffic with respect to the time trend
time_trend_graph_car <- cleaned_data %>%
  ggplot(aes(x = time_trend, y = daily_cars)) +
  geom_point(aes(color = daily_cars), alpha = 0.5) + 
  geom_smooth(method = "lm", se = FALSE, color = "blue") +  
  scale_color_gradient(low = "blue", high = "red") +
  labs(title = "Car Traffic Volume Over Time Trend",
       x = "Days Since Start",
       y = "Daily Car Traffic",
       color = "Daily Traffic") +
  theme_minimal()

# Display car traffic plot with time trend
time_trend_graph_car

```

The scatter plot illustrates the fluctuating daily car counts over time, with a trend line suggesting an initial increase in traffic followed by a decline.

# Model

This study uses negative binomial regression to address the problem of over-dispersion, which occurs when observed variance is greater than the mean and is frequently seen in count data. The dependent variable in the model is the daily count of cars, which is assumed to be impacted by temporal variations and the counts of alternative transportation modes (such as buses, bicycles, and pedestrians).

As such, we estimate the following model:

```{=tex}
\begin{align*} 
\log(\text{Count\_Cars}) &= \beta_0 + \beta_1 \times \text{Bus} + \beta_2 \times \text{Pedestrians} + \beta_3 \times \text{Bike} + \text{time\_trend} \\
\text{where} & \\
\log(\text{Count\_Cars}) & \text{ is the natural logarithm of the count of cars,} \\
\beta_0 & \text{ is the intercept of the model,} \\
\beta_1, \beta_2, \beta_3 & \text{ are the coefficients for the counts of buses, pedestrians, and bicycles,} \\
\text{time\_trend} & \text{ is the variable capturing the longitudinal changes.}
\end{align*}
```
## Model justification

There are two reasons for using negative binomial regression. First, since events tend to cluster together in both place and time, traffic count data usually show overdispersion. For such data, the negative binomial model makes sense since it allows the variance to surpass the mean. Second, given that traffic counts frequently deviate from a normal distribution, this model can account for their skewed nature.

The analysis is divided into two temporal models in order to evaluate the effects of the Route 508 Lake Shore streetcar service upgrades: `model_before`, which is the model before the improvements, and `model_after`, which is the model after the improvements. This method clarifies the relationship that existed before and after the policy intervention between the independent and dependent variables.

Together with a temporal trend, the `model_before` captures the baseline relationship between automobile traffic and other modes of transportation. As an illustration of possible changes in traffic patterns brought about by the improved streetcar service, the `model_after`, on the other hand, depicts the altered dynamics following the enhancement. By use of comparative analysis, these models demonstrate the effectiveness of the Route 508 enhancements in mitigating car congestion.

# Results

```{r residuals-before-model, fig.cap="Residuals of the negative binomial regression model before and after the transit improvement. This plot helps to visualize if there are any patterns or systematic deviations unexplained by the model.", echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
# Read and prepare data
analysis_data <- read.csv(here("data", "analysis_data", "cleaned-data.csv"))

# Scaling predictor variables for analysis
analysis_data <- analysis_data %>%
  mutate(
    daily_bus = scale(daily_bus),
    daily_peds = scale(daily_peds),
    daily_bike = scale(daily_bike)
  )


# Splitting the dataset into before and after a specific date for comparison
start_before <- as.Date("2011-12-31")
end_before <- as.Date("2014-12-31")
start_after <- as.Date("2018-11-01")
end_after <- as.Date("2020-02-29")

before_improvement <- subset(analysis_data, count_date >= start_before & count_date <= end_before)
after_improvement <- subset(analysis_data, count_date >= start_after & count_date <= end_after)

### Model data ####

# Building negative binomial regression models for before and after improvement
model_before <- glm.nb(daily_cars ~ daily_bus + daily_peds + daily_bike + time_trend, data = before_improvement)

model_after <- glm.nb(daily_cars ~ daily_bus + daily_peds + daily_bike + time_trend, data = after_improvement)

# Visualizing residuals to identify patterns
par(mfrow = c(1, 2))
plot(residuals(model_before), type = "p", main = "Residuals of Model Before Improvement")
plot(residuals(model_after), type = "p", main = "Residuals of Model After Improvement")

# Examining the normality of residuals
# hist(resid(model_before), main = "Histogram of Residuals - Before")
# hist(resid(model_after), main = "Histogram of Residuals - After")
```

Figures 6 and 7 show an examination of residuals from the negative binomial regression models before and after the public transit improvement. For the model before improvements, the residuals plot does not indicate any apparent patterns or systemic deviations, suggesting that the model's assumptions may be adequate for the data. After the transit improvements, the residuals' plot again shows no obvious patterns, which suggests a random scatter.

\newpage

```{r model summaries, echo=FALSE, message=FALSE, warning=FALSE}

# Read and prepare data
analysis_data <- read.csv(here("data", "analysis_data", "cleaned-data.csv"))
analysis_data$count_date <- as.Date(analysis_data$count_date)

# Scaling predictor variables for analysis
analysis_data <- analysis_data %>%
  mutate(
    daily_bus = scale(daily_bus, center = TRUE, scale = TRUE),
    daily_peds = scale(daily_peds, center = TRUE, scale = TRUE),
    daily_bike = scale(daily_bike, center = TRUE, scale = TRUE)
  )

# Splitting the dataset into before and after a specific date for comparison
start_before <- as.Date("2011-12-31")
end_before <- as.Date("2014-12-31")
start_after <- as.Date("2018-11-01")
end_after <- as.Date("2020-02-29")

before_improvement <- subset(analysis_data, count_date >= start_before & count_date <= end_before)
after_improvement <- subset(analysis_data, count_date >= start_after & count_date <= end_after)

analysis_data <- analysis_data %>%
  mutate(time_trend = as.numeric(count_date - min(count_date, na.rm = TRUE)))

# Building negative binomial regression models for before and after improvement
model_before <- glm.nb(daily_cars ~ daily_bus + daily_peds + daily_bike + time_trend, data = before_improvement)
model_after <- glm.nb(daily_cars ~ daily_bus + daily_peds + daily_bike + time_trend, data = after_improvement)

# Tidying the model summaries 
tidy_model_before <- tidy(model_before)
tidy_model_after <- tidy(model_after)

# Model before improvement
kable(tidy_model_before, format = "latex", booktabs = TRUE, caption = "Negative Binomial Regression Model - Before Improvement") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = F) %>%
  column_spec(1, bold = TRUE, width = "3cm") %>%
  column_spec(2, bold = TRUE, color = "blue", width = "3cm") %>%
  column_spec(3, width = "3cm")

kable(tidy_model_after, format = "latex", booktabs = TRUE, caption = "Negative Binomial Regression Model - After Improvement") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = F) %>%
  column_spec(1, bold = TRUE, width = "3cm") %>%
  column_spec(2, bold = TRUE, color = "blue", width = "3cm") %>%
  column_spec(3, width = "3cm")


```

Table 2 represents the traffic patterns before the improvement. These results produced significant coefficients for both the intercept and daily bus traffic, as well as the time trend variable. The intercept, representing the baseline log-count of cars when all other variables are zero, was significant with an estimate of 9.340 and a standard error of 0.072, indicating a high baseline traffic volume. Daily bus traffic had a positive and significant relationship with car traffic, with a coefficient of 0.261, suggesting that increases in bus traffic were associated with increases in car traffic. The time trend also showed a significant, albeit small, positive effect on car traffic `(β = 0.000287)`, indicating a gradual increase in car traffic over time.

Pedestrian and bicycle traffic, however, were not significantly associated with car traffic before the improvements (p \> 0.05 for both).

Post-improvement, the model exhibits significant changes in the dynamics of traffic patterns. The intercept remains significant but increases to 12.283, reflecting an overall higher baseline of car traffic possibly due to other factors. The coefficient for daily bus traffic notably decreases to 0.096, indicating a diminished relationship between bus and car traffic after the public transit improvements.

The time trend shifts from a slight increase to a significant decrease `(β = -0.001033)`, reflecting a reversal in the car traffic trend, which now shows a significant decline over time post-improvement.

Variables for pedestrian and bicycle traffic remained non-significant, suggesting that these modes of transport did not have a substantial impact on car traffic volumes in either model.

The model results indicate a significant reduction in the relationship between bus traffic and car congestion following the transit improvements, as shown by the decrease in the coefficient of daily bus traffic from 0.261 to 0.096. This suggests that improvements in the public transit system might be effectively alleviating car congestion. Additionally, the significant negative time trend in the post-improvement model underscores a successful overall reduction in car traffic between 2012 to 2020.

# Discussion

## First discussion point {#sec-first-point}

The introduction of improvements in Toronto's Route 508 Lake Shore streetcar has demonstrated a tangible reduction in the correlation between bus traffic and car congestion. This aligns with global evidence from cities like Amsterdam, where extensive public transit systems have significantly minimized car usage, fostering more sustainable urban environments [Source here]. Such systems not only alleviate congestion but also reduce environmental impact by lowering emissions and promoting cleaner air. This study confirms the view that public transit enhancements are critical for achieving urban sustainability goals.

-   Study's results highlight that improved public transit correlates with decreased car congestion.
-   Talk about public transit and urban sustainability.
-   Provide examples from other countries e.g., Netherlands (maybe focus on Amsterdam).
    -   Adding any additional countries could be helpful.

## Second discussion point

[Source] Upgraded public transit systems can spur economic growth by enhancing accessibility to employment and reducing travel-related expenses. In cities with robust public transit, property values tend to rise, and commercial areas see increased activity, illustrating the broader economic benefits of accessible public transportation. [Source] This study's findings underscore the potential of public transit to drive urban development by reshaping cityscapes into more pedestrian- and transit-friendly environments, thereby attracting more investment and improving the overall quality of urban life.

-   Effects of improved public transit such as economic.
-   How does public transit influence development in urban areas.

## Third discussion point

Furthermore, improving public transit infrastructure serves as a powerful tool for enhancing social equity by providing reliable and affordable transportation options to diverse populations, including the disabled, elderly, and low-income individuals. [Source] highlights the critical role of public transit in ensuring that mobility and access to urban opportunities are not luxuries but basic services accessible to all. By enabling equal participation in urban life, enhanced public transit systems can break down barriers and foster more inclusive communities.

-   Public transit can enhance accessibility for disabled communities, minors, elderly, students.
-   Public transit as an equalizer for class. i.e., people with less wealth may not be able to afford a car so improved public transit gives them access to more spaces for a lower individual cost.

## Weaknesses and next steps

While this research provides valuable insights, it acknowledges several limitations that could be addressed in future studies. The absence of detailed demographic and socioeconomic data limits the ability to deeply understand the differential impacts of public transit improvements across various population segments. Furthermore, external factors such as weather conditions and broader social changes, including those prompted by global crises like the COVID-19 pandemic, were not considered but could significantly affect public transit usage and perceptions.

The study's primary weakness lies in its narrow focus on traffic volume data without incorporating broader contextual factors that might influence these volumes. 

Future research should expand the scope to include a comparative analysis of multiple cities with varying transit modes to identify best practices and related influences. It would also be beneficial to integrate demographic data to explore how different groups are uniquely affected by public transit changes. 

-   Could improve if there is access to population data (maybe weather patterns).

-   Could benefit from a comparison with different cities and modes of transit.

-   Consider COVID, dangers associated with public transit and how that can affect this research which can affect attitude to public transit.

\newpage

\appendix

# Appendix {.unnumbered}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

## Diagnostics

\newpage

# References
